---
title: "Political Agency and Accountability Model."
---

## Simple Model : Framework.

```{r}
#| include: false
library(rethinking)
library(tidyverse)
library(conflicted)
```

I am trying to make a simple model to see various implications of changing important parameters of the political agency and accountablity model.

Let's take our model one step at a time.

The first variable is that of the 'state of the world'. This corresponds to some economic, political and/or sociological state of the world that voters and candidates/politicians correspond to. For simplicity, we are going to make this binary. Let's run the simulation of various states of the world 200 times with equal probability of each state ${0,1}$

```{r}
N <- 200
s <- sample(c(0,1), N, replace = TRUE)
head(s)
```

The model also assumes that there are essentially two candidate *types* : Congruent and Dissonant. A congruent politician acts/formulates policies in accordance with the state of the world while a dissonant politician acts the exact opposite (maybe because she has been bought by large business through some rent, or maybe she just loves chaos). Let $\pi$ be the proportion of congruent candidates in the candidate population.

```{r}
pi <- 0.5
candpool <- rbinom(N, 1, prob = pi) + 1 
dens(rnorm(N))
# here congruent is 1 and dissonant is 2.
```
